---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- read10xCounts(samples = c("pbmc5k" ="/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix"), col.names = T)


  
  

sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
sce <- sce[, colSums(assay(sce, "counts")) > 0]
sce
```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(
  sce,
  lower = 100,
  niters = 10000,
  assay.type = "counts")
out #the column we're most interested in is FDR (the multiple testing correction)
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
positions <- which(out$FDR < 0.01)

sce_filtered_emptydrops <- sce[, positions]
sce_filtered_emptydrops
remove(positions)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
plot_data <- tibble(library_size = colSums(assay(sce, "counts")), droplet_status = out$FDR < 0.01)
plot_data 
  

ggplot(plot_data, aes(x=droplet_status, y=library_size)) +
  geom_violin() + scale_y_log10()

#NA: what we set as empty (< 100) 
#False: what it identified as empty 
#True: what it identified as non-empty
  
  
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  


all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops)
colData(sce_scdblfinder)
table(sce_scdblfinder$scDblFinder.class)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer: Droplets with scDblFinder.class of doublet 
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
is_singlet <- sce_scdblfinder$scDblFinder.class == "singlet"
sce_filtered_scdblfinder <- sce_scdblfinder[, is_singlet]
sce_filtered_scdblfinder
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  library_size = colSums(counts(sce_scdblfinder)), 
  droplet_type = sce_scdblfinder$scDblFinder.class)
plot_data

ggplot(plot_data) + geom_violin(aes(x=droplet_type, y = library_size))
  
  
  
```

```{r}
is.mt <- grep(rowData(sce_scdblfinder)$Symbol, pattern = "^MT-")
percent.mt <- colSums(counts(sce_scdblfinder)[is.mt, ]) / colSums(counts(sce_scdblfinder)) * 100
plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  mt_percent = percent.mt, 
  droplet_type = sce_scdblfinder$scDblFinder.class)  
ggplot(plot_data) +
  geom_violin(aes(x = droplet_type, y = mt_percent), fill = "orange") +
  theme_cowplot()
  
  
  
```

```{r}
set.seed(1)
sce_scdblfinder <- logNormCounts(sce_scdblfinder) 
sce_scdblfinder <- runPCA(sce_scdblfinder)

plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  PC1 = reducedDim(sce_scdblfinder, "PCA")[, "PC1"],
  PC2 = reducedDim(sce_scdblfinder, "PCA")[, "PC2"],
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot() +
  geom_point(data = plot_data %>% filter(droplet_type == "singlet"),
             mapping = aes(x = PC1, y = PC2),
             size = 0.5,
             colour = "black") +
  geom_point(data = plot_data %>% filter(droplet_type == "doublet"),
             mapping = aes(x = PC1, y = PC2),
             size = 0.5,
             colour = "red") +
  theme_cowplot()

set.seed(1)
sce_scdblfinder <- runUMAP(sce_scdblfinder, dimred = "PCA", n_dimred = 20)

plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  UMAP1 = reducedDim(sce_scdblfinder, "UMAP")[, 1],
  UMAP2 = reducedDim(sce_scdblfinder, "UMAP")[, 2],
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot() +
  geom_point(data = plot_data %>% filter(droplet_type == "singlet"),
             mapping = aes(x = UMAP1, y = UMAP2),
             size = 0.5,
             colour = "black") +
  geom_point(data = plot_data %>% filter(droplet_type == "doublet"),
             mapping = aes(x = UMAP1, y = UMAP2),
             size = 0.5,
             colour = "red") +
  theme_cowplot()
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_scdblfinder)
colData(sce_decontX)

```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
colData(sce_decontX)
umap <- plotDimReduceCluster(colData(sce_decontX)$decontX_clusters, #this is an interger vector of cell cluster labels
                             dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
                             dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
                             )
umap
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}

rowData(sce_decontX)

rownames(sce_decontX) <- uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)

plotDimReduceFeature(decontXcounts(sce_decontX), #logcounts() will show the contaminated data
                     features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"),
                     dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
                     dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2] 
                        )
                     
  
  
  
  
  
  

```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list(
  Tcells = c("CD3D", "CD3E"),
  Monocytes = c("LYZ", "S100A8", "S100A9"),
  Bcells = c("CD79A", "CD79B", "MS4A1"),
  NKcells = "GNLY",
  Megakaryocytes = "PPBP"
  )
  
  
cellTypeMappings <- list(
  Tcells = 1,
  Monocytes = 2,
  Bcells = 3,
  NKcells = 5,
  Megakaryocytes = 7
)

plotDecontXMarkerPercentage(sce_decontX, 
                            markers = markers,
                            groupClusters = cellTypeMappings,
                            assayName = c("counts", "decontXcounts")
                      
)
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r, fig.width=10, fig.height=10}

sce_decontX <- logNormCounts(sce_decontX, name="decontXlogcounts", assay.type="decontXcounts") 
assays(sce_decontX) 

#Log-normalise 
plotDecontXMarkerExpression(sce_decontX,
                            markers = markers,
                            groupClusters = cellTypeMappings,
                            assayName = c("logcounts", "decontXlogcounts"))                        
  

#Alternatively, if you feel log transformation is enough 
plotDecontXMarkerExpression(sce_decontX,
                            markers = markers,
                            groupClusters = cellTypeMappings,
                            assayName = c("counts", "decontXcounts"),
                            log1p = T)

```
